A simplified dynamic bioenergetic model for coral bleaching

### 2016-10-18 JTB_ms/JTB_ms.tex
text updates

NPQ capacity of symbiont, relative to C-fixation

@@

@@

\(k\)
Exponent on ROS production rate

@@

The symbiont produces fixed carbon through photosynthesis, a process
represented here by a single SU with two substrates: light (photons) and
inorganic carbon (CO\textsubscript{2}). The amount of light absorbed by
the symbiont depends on the scalar irradiance at the site of light
absorption, which is modified substantially relative to external
downwelling irradiance owing to multiple scattering by the coral
skeleton and self-shading by surrounding symbionts (Enríquez, Méndez,
and Iglesias-Prieto 2005; Marcelino et al. 2013). We used data from
Marcelino et al. (2013) to empirically derive an ``amplification
factor'' \(A\) (the ratio of internal scalar irradiance to external
downwelling irradiance) as a function of symbiont density (expressed as
symbiont to host biomass ratio), which is specified as:

@@

This amplification factor is then multiplied by the external downwelling
irradiance \(L\) and the effective light-absorbing surface area of
symbiont biomass \(\bar{a}^*\) to specify the total amount of light
absorbed:

@@

We then specify the input of CO\textsubscript{2} to photosynthesis as
\(j_{CO_2}\), which is mediated by the host. This encompasses
potentially diverse mechanisms by which the host may enhance
CO\textsubscript{2} availability for photosynthesis (i.e.,
carbon-concentrating mechanisms (CCMs)), including active transport of
bicarbonate, carbonic anhydrase-catalyzed conversion of bicarbonate to
CO\textsubscript{2} to promote diffusion toward the symbiont (Tansik,
Fitt, and Hopkinson 2015), and acidification of the symbiosome to
increase localized CO\textsubscript{2} concentrations around the
symbiont (Barott et al. 2014). Since these mechanisms require energetic
input by the host, the magnitude of CO\textsubscript{2} delivery is set
proportional to \(j_{eC}\), the amount of excess carbon from the host
biomass SU, assuming that some of this carbon is respired to energize
these CCMs. This formulation means that the symbiont indirectly ensures
its own CO\textsubscript{2} supply by providing the host with fixed
carbon (Wooldridge 2013), which establishes a positive feedback in the
model that proves to be essential in simulating the phenomenon of coral
bleaching. The parameter \(k_{CO_2}\) is included to scale the efficacy
of host CCMs, i.e., how many moles of CO\textsubscript{2} can be
delivered from the environment to the symbiont per mole of fixed carbon.
Adjusting this parameter offers the user the opportunity to compare
different rates of CO\textsubscript{2} delivery that may characterize
different coral species (Wooldridge 2014a). The input of
CO\textsubscript{2} to the photosynthesis SU is therefore specified as:

@@

where \(j_{CPm}\) is the maximum specific rate of photosynthesis, and
\(c_{ROS}\) is the relative rate of reactive oxygen species production
(see below). Dividing the photosynthetic rate by \(c_{ROS}\) causes a
decline in response to photooxidative stress (i.e., photoinhibition).

@@

If light energy further exceeds the total capacity of both carbon
fixation and NPQ, then reactive oxygen species (ROS) are produced. We
represent this as a relative quantity \(c_{ROS}\), which takes a value
of 1 when all light absorbed can be quenched by photochemistry and NPQ,
and increases as the amount of excess excitation energy further
increases:

\begin{equation} c_{ROS} = 1 + \bigg({{(j_{eL} - j_{NPQ})_+} \over k_{ROS}}\bigg)^k \end{equation}

where \(k_{NPQ}\), \(k_{ROS}\), and \(k\) are parameters of the symbiont
that determine the onset and rate of ROS production. Importantly,
\(c_{ROS}\) as specified here is not a function of absolute light
absorption, but rather the amount of excess light energy, \(j_{eL}\),
after accounting for carbon fixation and NPQ. A direct consequence of
this formulation is that CO\textsubscript{2}-limitation of
photosynthesis can lead to ROS production, an important mechanism
(Wooldridge 2009) that was not captured by previous representations of
photooxidative stress (Eynaud, Nisbet, and Muller 2011). Moreover, this
formulation allows functional diversity among symbiont types to be
explored by changing the parameters \(k_{NPQ}\), \(k_{ROS}\), and \(k\).

@@

This second component of symbiont biomass loss can represent both
photodamage and/or symbiont expulsion (i.e., bleaching), both of which
occur in response to high levels or ROS production. The parameter \(b\)
is included to scale biomass loss due to bleaching in response to ROS.
(Note that recycling of symbiont biomass turnover (\(rNS\) and \(rCS\))
only occurs based on the maintenance component of turnover (i.e.,
\(j_{ST}^0\)), and not the bleaching-related component, as this loss
represents biomass that is damaged or expelled from the host).

@@

A time-stepping Euler method was used to solve the state equations since
the production and rejection fluxes of the SUs are imiplicitly defined.
Specifically, the rejection fluxes of carbon and nitrogen from the
symbiont and host biomass SUs act as reciprocal input fluxes to the
other SU. In addition, the rejection flux of excitation energy from the
photosynthesis SU acts to reduce its own production flux (i.e.,
photoinhibition), and hence a discretized time-stepping procedure was
necessary. A vector of time values was created for each simulation run,
along which dynamic environmental forcing functions (irradiance, DIN,
and prey abundance) can be designed. These vectors, along with initial
values of symbiont and host biomass, then serve as input to the
time-stepping function, which solves for the current system state using
values of the previous system state where necessary. A time step of 0.1
days was used for all simulations, which were performed using R code
that is available in the data repository accompanying this article:
\url{github.com/jrcunning/Rcoral}.

@@

where \(j_{S1}\) and \(j_{S2}\) are the specific input fluxes of the two
substrates and \(j_{Pm}\) is the maximum specific product formation
rate, all with units of Cmol Cmol\textsuperscript{-1}
d\textsuperscript{-1}.

@@

simulations in constant environs show either positive or negative
exponential growth. some environments show two steady states, one with
positive and one with negative growth, depending on initial conditions.
based on numerical analysis, we see ultimate exponential growth or
decline. the mechanism that produces these alternate steady states is:
\ldots{}. . for practical purposes, we are interested in the positive
growth steady states and their properties and how they change across
environmental gradients (see section X). but we are also interested in
the possibility for flips to the alternate steady states (see section
Y).

@@

\emph{emphasize that steady state is exponential growth, and fluxes are
per unit biomass/specific rates} To analyze qualitative behavior, we ran
the model to steady state across gradients of external irradiance and
nutrients (Fig. 2), which revealed patterns consistent with observed
phenomena in corals. Predicted growth rates are low at low light and
DIN, and begin increasing as both of these factors increase (Fig. 2A).
Low light is limiting to photosynthesis (Fig. 2D), which leads to
carbon-limitation of host growth (Fig. 2B) and an associated increase in
the symbiont to host biomass ratio (Fig. 2C). In agreement with this
trend are many observations of negative correlation between irradiance
and symbiont density (Stimson 1997; Brown et al. 1999; Fitt et al. 2000;
Titlyanov et al. 2001). As higher light alleviates light-limitation of
photosynthesis, growth becomes less carbon-limited. Similarly, elevating
DIN up to \textasciitilde{}2 µM alleviates nitrogen-limitation (Fig.
2B). Increased growth at higher DIN is predicted by the DEB model of
Muller et al. (2009), and has also been observed experimentally
(Muller-Parker et al. 1994; Tanaka et al. 2007; Tanaka et al. 2013).
However, DIN elevation beyond a certain point (e.g., 2-3 µM in these
simulations) has little effect on growth as carbon becomes limiting
(Fig. 2B). Very high nutrient levels may even reduce growth (Shantz,
Lemoine, and Burkepile 2015), although these impacts are not likely to
occur within the range of concentrations considered here (\textless{}4
µM) (Ferrier-Pagès et al. 2000).

@@

\subsection{varying light: consequences of excess light and
light-limitation}\label{varying-light-consequences-of-excess-light-and-light-limitation}

\subsection{varying DIN: role of N-limitation and consequences of
elevated
N}\label{varying-din-role-of-n-limitation-and-consequences-of-elevated-n}

\subsection{elevated nutrients cause system to become C-limited (photo:
CO2 limited; host: fixed-C
limited)}\label{elevated-nutrients-cause-system-to-become-c-limited-photo-co2-limited-host-fixed-c-limited}

\subsection{C-limitation increases susceptibility to
bleaching}\label{c-limitation-increases-susceptibility-to-bleaching}

\subsection{consequences of a less efficient but more tolerant
symbiont}\label{consequences-of-a-less-efficient-but-more-tolerant-symbiont}

\subsection{Rate of symbiont repopulation (Berner et al.
1993)}\label{rate-of-symbiont-repopulation-berner-et-al.-1993}

\subsection{Bicarbonate addition promotes coral growth (Marubini and
Thake
1999)}\label{bicarbonate-addition-promotes-coral-growth-marubini-and-thake-1999}

\subsection{Seasonal change in sym. density is larger when nutrients are
elevated (Stimson 1997). Such a phenomenon was predicted by Fitt et al.
2000.}\label{seasonal-change-in-sym.-density-is-larger-when-nutrients-are-elevated-stimson-1997.-such-a-phenomenon-was-predicted-by-fitt-et-al.-2000.}

\subsection{Death rule: negative growth for a certain amount of time
should be
defensible.}\label{death-rule-negative-growth-for-a-certain-amount-of-time-should-be-defensible.}

>>>>

NPQ capacity of symbiont

@@

112

@@

@@

The symbiont produces fixed carbon through photosynthesis, a process
represented here by a single SU with two substrates: light (photons) and
inorganic carbon (CO\textsubscript{2}). The amount of light absorbed by
the symbiont depends on the scalar irradiance at the site of light
absorption, which is modified substantially relative to external
downwelling irradiance owing to multiple scattering by the coral
skeleton and self-shading by surrounding symbionts (Enríquez, Méndez,
and Iglesias-Prieto 2005; Marcelino et al. 2013). We used data from
Marcelino et al. (2013) to empirically derive an amplification factor,
\(A\), indicating the ratio of internal scalar irradiance to external
downwelling irradiance as a function of symbiont density (S:H biomass),
which is specified as:

@@

This amplification factor is then multiplied by the external downwelling
irradiance \(L\) and the effective light-absorbing surface area of
symbiont biomass \(\bar{a}^*\) to specify the total light absorption:

@@

We then specify the input of CO\textsubscript{2} to photosynthesis as
\(j_{CO_2}\), a flux mediated by the host that encompasses potentially
diverse carbon-concentrating mechanisms (CCMs), including active
transport of bicarbonate, carbonic anhydrase-catalyzed conversion of
bicarbonate to CO\textsubscript{2} to promote diffusion toward the
symbiont (Tansik, Fitt, and Hopkinson 2015), and acidification of the
symbiosome to increase localized CO\textsubscript{2} concentrations
around the symbiont (Barott et al. 2014). Since these mechanisms require
energetic input by the host, the magnitude of CO\textsubscript{2}
delivery is proportional to the host's excess carbon, \(j_{eC}\),
assuming that some of this carbon is respired to energize these CCMs.
This formulation means that the symbiont indirectly ensures its own
CO\textsubscript{2} supply by providing fixed carbon to the host
(Wooldridge 2013), which establishes a positive feedback in the model
that leads to the existence of multiple stable states, which we discuss
below in the context of coral bleaching. The parameter \(k_{CO_2}\)
scales the efficacy of host CCMs, which enables the comparison of
different rates of CO\textsubscript{2} delivery that may characterize
different coral species (Wooldridge 2014a). The input of
CO\textsubscript{2} to the photosynthesis SU is therefore specified as:

@@

where \(j_{CPm}\) is the maximum specific rate of photosynthesis, and
\(c_{ROS}\) is the relative rate of reactive oxygen species production
(see below). Dividing the photosynthetic rate by \(c_{ROS}\) causes a
decline in response to photooxidative stress at high light levels, and
the emergent outcome of this SU formulation demonstrates a classic
photoinhibition response (see Fig. S1).

@@

If light energy further exceeds the capacity of both photochemistry and
NPQ, then reactive oxygen species (ROS) are produced. We represent this
as a relative quantity \(c_{ROS}\), taking a value of 1 when all light
energy is quenched by photochemistry and NPQ, and increasing as the
amount of excess excitation energy increases:

\begin{equation} c_{ROS} = 1 + {{(j_{eL} - j_{NPQ})_+} \over k_{ROS}} \end{equation}

where \(k_{ROS}\) is a parameter of the symbiont that determines the
rate of ROS production (specifically, the amount of excess excitation
energy that doubles ROS production relative to baseline levels).
Importantly, \(c_{ROS}\) is specified here not as a function of absolute
external light, but rather the amount of excess light, \(j_{eL}\), after
accounting for quenching by carbon fixation and NPQ. A direct
consequence of this formulation is that CO\textsubscript{2}-limitation
of photosynthesis can lead to ROS production, an important mechanism
(Wooldridge 2009) that was not captured by previous representations of
photooxidative stress (Eynaud, Nisbet, and Muller 2011).

@@

This second component of symbiont biomass loss represents both
photodamage and/or symbiont expulsion (i.e., bleaching), both of which
occur in response to high levels of ROS production. The parameter \(b\)
is included to scale biomass loss due to bleaching in response to ROS.
(Note that recycling of symbiont biomass turnover (\(rNS\) and \(rCS\))
only occurs based on the maintenance component of turnover (i.e.,
\(j_{ST}^0\)), and not the bleaching-related component, as this loss
represents biomass that is damaged or expelled from the host.)

@@

A time-stepping Euler method was used to solve the state equations since
the production and rejection fluxes of the SUs are imiplicitly defined
due to feedbacks present in the model. Specifically, the rejection
fluxes of carbon and nitrogen from the symbiont and host biomass SUs act
as reciprocal input fluxes to the other SU. In addition, the
photosynthesis SU receives CO\textsubscript{2} proportional to the
carbon rejection flux from the host biomass SU, and the rejection flux
of excitation energy from the photosynthesis SU acts to reduce its own
production (i.e., photoinhibition). A vector of time steps is used for
each simulation, along which dynamic environmental forcing functions
(irradiance, DIN, and prey abundance) can be designed. These vectors,
along with initial values of symbiont and host biomass, then serve as
input to the time-stepping function, which solves for the current system
state using values of the previous system state where necessary. A time
step of 0.1 days was used for all simulations, which were performed
using R code that is available in the data repository accompanying this
article: \url{github.com/jrcunning/Rcoral}.

@@

where \(j_{S1}\) and \(j_{S2}\) are the specific input fluxes of the two
substrates and \(j_{Pm}\) is the maximum specific product formation
rate, in units of Cmol Cmol\textsuperscript{-1} d\textsuperscript{-1}.

@@

In a constant environment, the system ultimately reaches a steady state
of exponential growth or decline. Under some conditions, alternate
steady states can be reached depending on initial values of symbiont and
host biomass. The mechanism that produces these alternate steady states
is the positive feedback between carbon-limitation of the host and
CO\textsubscript{2}-limitation of photosynthesis. If the host becomes
carbon-limited, the efficacy of its CCMs are reduced and photosynthesis
becomes progressively more CO\textsubscript{2}-limited. In turn, less
photosynthate is translocated to the host, causing further
carbon-limitation and an ultimate decline in biomass; this steady state
represents a dysfunctional symbiosis. However, if the host is not
carbon-limited, it effectively delivers CO\textsubscript{2} to
photosynthesis and ensures its continued receipt of fixed carbon from
the symbiont; this system is nitrogen-limited and represents a
functional symbiosis with positive exponential growth. Either of these
stable states may be reached under (some sets of) constant environmental
conditions, depending on the initial values of symbiont and host
biomass: if symbiont biomass is initially very low (i.e., a ``bleached''
coral), the system cannot escape the carbon-limited state and cannot
grow. However, if symbiont biomass is initially high (i.e., a healthy
coral), then the system maintains positive growth in a nitrogen-limited
state. For practical purposes, we consider only positive growth steady
states in the ``steady state behavior'' section below. In the subsequent
``dynamic behavior'' section, we explore more thoroughly the conditions
that cause the system to switch between alternate stable states, and
interpret these dynamics in the context of coral bleaching.

@@

To analyze qualitative behavior, we ran the model to steady state across
gradients of external irradiance and nutrients (Fig. 2), which revealed
patterns consistent with observed phenomena in corals. Predicted growth
rates are low at low light and DIN, and begin increasing as both of
these factors increase (Fig. 2A). Low light is limiting to
photosynthesis (Fig. 2D), which leads to carbon-limitation of host
growth (Fig. 2B) and an associated increase in the symbiont to host
biomass ratio (Fig. 2C). In agreement with this trend are many
observations of negative correlation between irradiance and symbiont
density (Stimson 1997; Brown et al. 1999; Fitt et al. 2000; Titlyanov et
al. 2001). As higher light alleviates light-limitation of
photosynthesis, growth becomes less carbon-limited. Similarly, elevating
DIN up to \textasciitilde{}2 µM alleviates nitrogen-limitation (Fig.
2B). Increased growth at higher DIN is predicted by the DEB model of
Muller et al. (2009), and has also been observed experimentally
(Muller-Parker et al. 1994; Tanaka et al. 2007; Tanaka et al. 2013).
However, DIN elevation beyond a certain point (e.g., 2-3 µM in these
simulations) has little effect on growth as carbon becomes limiting
(Fig. 2B). Very high nutrient levels may even reduce growth (Shantz,
Lemoine, and Burkepile 2015), although these impacts are not likely to
occur within the range of concentrations considered here (\textless{}4
µM) (Ferrier-Pagès et al. 2000).

@@

Therefore, we are highly interested in the conditions under which the
system switches from one state to another, representing the conditions
under which corals ``bleach'' due to progressive, reinforcing
carbon-limitation. Interestingly, this state can be reached in multiple
ways, by the acceleration of photoinhibition and photodamage, or by high
external DIN, which causes the symbiont to share less carbon,
subsequently causing its own CO2-limitation, and progressive
C-limitation of the system. \#\# varying light: consequences of excess
light and light-limitation \#\# varying DIN: role of N-limitation and
consequences of elevated N \#\# elevated nutrients cause system to
become C-limited (photo: CO2 limited; host: fixed-C limited) \#\#
C-limitation increases susceptibility to bleaching \#\# consequences of
a less efficient but more tolerant symbiont \#\# Rate of symbiont
repopulation (Berner et al. 1993) \#\# Bicarbonate addition promotes
coral growth (Marubini and Thake 1999) \#\# Seasonal change in sym.
density is larger when nutrients are elevated (Stimson 1997). Such a
phenomenon was predicted by Fitt et al. 2000. \#\# Death rule: negative
growth for a certain amount of time should be defensible.
